A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-11-06 Draft14.tex
small changes

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{\today}
\author{Ashia C. Wilson \and Benjamin Recht \and Michael I. Jordan}
\date{\today}
\maketitle
\begin{abstract}
Momentum methods play a central role in optimization. Several momentum methods are provably optimal, and all use a technique called {\em estimate sequences} to analyze their convergence properties.  The technique of estimate sequences has long been considered difficult to understand, leading many researchers to generate alternative, ``more intuitive'' methods and analyses. In this paper we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simple and unified analysis of many existing momentum algorithms, introduce several new algorithms, and strengthen the connection between algorithms and continuous-time dynamical systems. 
\end{abstract}

@@

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratic costs, Polyak was able to provide an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's Iterative Method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex functions. Indeed, Lessard~\emph{et al.} derived a simple one-dimensional counterexample where the standard Heavy Ball method does not converge~\cite{Lessard14}.

@@

Until recently, this ODE perspective on optimization algorithms was largely abandoned in favor of Nesterov's estimate sequence framework.  However, there has been a recent resurgence in interest in ODE-based analyses.  Su, Boyd and Cand\`es~\cite{SuBoydCandes14}, showed that the continuous-time limit of Nesterov's accelerated gradient descent method corresponds to the following second-order ODE, 
 \begin{equation}\label{Eq:SuBoydCandes}
 \ddot X_t + \frac{r}{t} \dot X_t + \nabla f(X_t) = 0,\,\,\,\,r \geq 3.
 \end{equation}
This is equivalent to~\eqref{eq:heavy-ball-ode} for an object with unit mass, where now the time-dependent damping term $\gamma_t = r\log t$ is logarithmic.  They also established a convergence rate of $O(1/t^2)$ for this continuous-time system via a Lyapunov function argument. This matches the $O(1/\epsilon k^2)$ convergence rate of accelerated gradient descent, given the discretization $t= \sqrt\epsilon k$. Thus, the convergence rate is maintained in passing from continuous time to discrete time. Two things remain unclear from their analysis however:  they introduce an additional sequence $\{y_k\}_{k=1}^\infty$ while discretizing the ODE~\cite[Sec 2]{Acceleration},  which does not correspond to a straightforward discretization technique; and they do not demonstrate what connection---if any---there is between the Lypaunov argument used to analyze the continuous-time dynamics and the technique of estimate sequences used to analyze the discrete-time algorithm.

@@

Note that in both the strongly convex and weakly convex setting, the structure of the Lyapunov function allows us to derive a rate of convergence for the optimality gap $f(X_t) - f(x^\ast)$. In particular, the Lyapunov property $\E_t \leq \E_0$ implies 
\begin{equation}\label{eq:Guarantee1}
f(X_t) - f(x^\ast) \leq \frac{e^{\beta_0}(f(X_0) - f(x^\ast)) + D_h(x^\ast, X_0 + e^{-\alpha_0} \dot X_0)}{e^{\beta_t}}
\end{equation}
and 
\begin{equation}
f(X_t) - f(x^\ast) \leq \frac{e^{\beta_0}(f(X_0) - f(x^\ast) +\mu D_h(x^\ast ,X_0 -e^{-\alpha_0} \dot X_0))}{e^{\beta_t}},
\end{equation}
for~\eqref{Eq:E} and \eqref{eq:general-sc-lyap} respectively. The typical convention is that we start from rest $\dot X_0 = 0$. Notice that we obtain the same rates of convergence in these two settings. However, the family of dynamics corresponding to the Lyapunov function in the strongly convex setting uses the strong convexity of $f$, and this allows us to obtain a tighter bound in discrete time.

\section{Momentum Methods}
\label{Sec:MomLyap}
In this section, we introduce two families of momentum dynamics, both which are generated by functionals defined on continuous-time curves. We illustrate how to derive a Lyapunov function from these dynamics. Next, we show how these dynamics can be discretized to obtain optimization algorithms. The same discrete-time mappings can be applied to the Lyapunov function, resulting in a tool which can be used to analyze the discrete-time algorithms. We provide examples of this procedure for the quasi-monotone subgradient algorithm, accelerated gradient descent, the conditional gradient algorithm, and several other novel algorithms. 
 \subsection{Lyapunov Analysis}
We study dynamics that are based on the Euler-Lagrange equation \eqref{Eq:ELBreg} in the setting where the ideal scaling~\eqref{Eq:IdeScaBet} holds with equality~$\dot \beta_t = e^{\alpha_t}$. Notice, now the entire Bregman Lagrangian is parameterized by $\beta_t$, which determines the convergence rate:
 \begin{subequations}\label{Eq:EL}
 \begin{align}
 Z_t &= X_t + \frac{1}{\dot \beta_t} \dot X_t \label{Eq:ELZ},\\
 \frac{d}{dt}\nabla h(Z_t) &= -\dot \beta_t e^{\beta_t} \nabla f(X_t)\label{Eq:ELH}.
 \end{align}
 \end{subequations}
We begin by demonstrating how to derive the Lyapunov function~\eqref{Eq:E} for the momentum dynamics~\eqref{Eq:ELBreg}; this derivation is similar in spirit to the Lyapunov analysis of mirror descent by Nemirovski and Yudin.

@@

\subsection{The Quasi-Monotone Subgradient Method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method.  It also achieves the $O(1/\sqrt{k})$ lower bound for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence-rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded} and show later how to extend the analysis to incorporate bounded subgradients. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit Euler method applied to~\eqref{Eq:Z} and the implicit Euler method applied to~\eqref{Eq:X} (where now, we take $e^{\beta_t} = A_{k+1}$, while leaving our identification for $\frac{d}{dt} e^{\beta_t}$ unchanged):

>>>>

\begin{document}
\title{A Lyapunov Analysis of Momentum Methods \\ in  Optimization}
%\title{A Lyapunov Perspective on Momentum Methods \\ in  Optimization}
\date{\today}
\author{Ashia C. Wilson \and Benjamin Recht \and Michael I. Jordan}
\date{\today}
\maketitle
\begin{abstract}
Momentum methods play a central role in optimization. Several momentum methods are provably optimal, and all use a technique called {\em estimate sequences} to analyze their convergence properties.  The technique of estimate sequences has long been considered difficult to understand, leading many researchers to generate alternative, ``more intuitive'' methods and analyses. In this paper we show there is an equivalence between the technique of estimate sequences and a family of Lyapunov functions in both continuous and discrete time. This framework allows us to develop a simple and unified analysis of many existing momentum algorithms, introduce several new algorithms, and most importantly, strengthen the connection between algorithms and continuous-time dynamical systems. 
\end{abstract}

@@

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratic costs, Polyak was able to provide an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's Iterative Method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex cost functions. Indeed, Lessard~\emph{et al.} derived a simple one-dimensional counterexample where the standard Heavy Ball method does not converge~\cite{Lessard14}.

@@

Until recently, this ODE perspective on optimization algorithms was largely abandoned in favor of Nesterov's estimate sequence framework.  However, there has been a recent resurgence in interest in ODE-based analyses.  Su, Boyd and Cand\`es~\cite{SuBoydCandes14}, showed that the continuous-time limit of Nesterov's accelerated gradient descent method corresponds to the following second-order ODE, 
 \begin{equation}\label{Eq:SuBoydCandes}
 \ddot X_t + \frac{r}{t} \dot X_t + \nabla f(X_t) = 0,\,\,\,\,r \geq 3.
 \end{equation}
This is equivalent to~\eqref{eq:heavy-ball-ode}, where the dynamics model an object with unit mass and the time-dependent damping term $\gamma_t = r\log t$ is logarithmic.  They also established a convergence rate of $O(1/t^2)$ for this continuous-time system via a Lyapunov function argument. This matches the $O(1/\epsilon k^2)$ convergence rate of accelerated gradient descent, given the discretization $t= \sqrt\epsilon k$. Thus, the convergence rate is maintained in passing from continuous time to discrete time. Two things remain unclear from their analysis however:  they introduce an additional sequence $\{y_k\}_{k=1}^\infty$ while discretizing the ODE~\cite[Sec 2]{Acceleration},  which does not correspond to a straightforward discretization technique; and they do not demonstrate what connection---if any---there is between the Lypaunov argument used to analyze the continuous-time dynamics and the technique of estimate sequences used to analyze the discrete-time algorithm.

@@

Note that in both the strongly convex and weakly convex setting, the structure of the Lyapunov function allows us to derive a rate of convergence for the optimality gap $f(X_t) - f(x^\ast)$. In particular, the Lyapunov property $\E_t \leq \E_0$ implies 
\begin{equation}\label{eq:Guarantee1}
f(X_t) - f(x^\ast) \leq \frac{e^{\beta_0}(f(X_0) - f(x^\ast)) + D_h(x^\ast, X_0 + e^{-\alpha_0} \dot X_0)}{e^{\beta_t}}
\end{equation}
and 
\begin{equation}
f(X_t) - f(x^\ast) \leq \frac{e^{\beta_0}(f(X_0) - f(x^\ast) +\mu D_h(x^\ast ,X_0 -e^{-\alpha_0} \dot X_0))}{e^{\beta_t}},
\end{equation}
for~\eqref{Eq:E} and \eqref{eq:general-sc-lyap} respectively. The typical convention is that the dynamics start from rest $\dot X_0 = 0$. Notice that we obtain the same rates of convergence in these two settings. However, the family of dynamics corresponding to the Lyapunov function in the strongly convex setting uses the strong convexity of $f$, and this allows us to obtain a tighter bound in discrete time.

\section{Momentum Methods}
\label{Sec:MomLyap}
In this section, we introduce two families of momentum dynamics, both which are generated by functionals defined on continuous-time curves. We illustrate how to derive a Lyapunov function from these dynamics. Next, we show how these dynamics can be discretized to obtain optimization algorithms. The same discrete-time mappings can be applied to the Lyapunov function, resulting in a tool which can be used to analyze the discrete-time algorithms. We provide examples of this procedure for the quasi-monotone subgradient algorithm, accelerated gradient descent, the conditional gradient algorithm, and several other novel algorithms. 
 \subsection{Lyapunov Analysis}
We study dynamics that are based on the Euler-Lagrange equation \eqref{Eq:ELBreg} in the setting where the ideal scaling~\eqref{Eq:IdeScaBet} holds with equality~$\dot \beta_t = e^{\alpha_t}$. Notice, now the entire Bregman Lagrangian is parameterized by $\beta_t$, which determines the convergence rate:
 \begin{subequations}\label{Eq:EL}
 \begin{align}
 Z_t &= X_t + \frac{1}{\dot \beta_t} \dot X_t \label{Eq:ELZ},\\
 \frac{d}{dt}\nabla h(Z_t) &= -\dot \beta_t e^{\beta_t} \nabla f(X_t)\label{Eq:ELH}.
 \end{align}
 \end{subequations}
We begin by demonstrating how to derive the Lyapunov function~\eqref{Eq:E} for the momentum dynamics~\eqref{Eq:EL}; this derivation is similar in spirit to the Lyapunov analysis of mirror descent by Nemirovski and Yudin.

@@

\subsection{The Quasi-Monotone Subgradient Method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method.  It also achieves the $O(1/\sqrt{k})$ lower bound for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence-rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded} and show later how the analysis extends to the setting where we only have access to bounded subgradients. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit Euler method applied to~\eqref{Eq:Z} and the implicit Euler method applied to~\eqref{Eq:X} (where now, we take $e^{\beta_t} = A_{k+1}$, while leaving our identification for $\frac{d}{dt} e^{\beta_t}$ unchanged):
